---
title: "Data wrangling - extra code"
subtitle: "BSTA 511/611 Fall 2025 Day 5, OHSU"
author: "Meike Niederhausen, PhD"
date: "10/15/2025"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: cerulean
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: "setup"
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, fig.height=3, fig.width=5, message = F)
```


```{r}
library(tidyverse)
library(janitor)
```

# Slides pdf

[Link](../slides_pdf/Day03_bsta511_part3_data_wrangling.pdf) to pdf of slides


# Goals for today

More data wrangling examples and tools


* Subsetting data
  * `filter`ing rows
  * `select`ing columns

* More wrangling for columns:
  * `relocate` columns
  * `rename` columns



* Creating new variables
  * `mutate`

* Reshaping data
  * wide vs long data
      * make wide data long
      * make long data wide


Check out the [data transformation cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) for many more data wrangling commands!



# Case study: discrimination in developmental disability support (1.7.1)

* In the US, individuals with developmental disabilities typically receive services and support from state governments. 
  * California allocates funds to developmentally disabled residents through the Department of Developmental Services (DDS)
  * Recipients of DDS funds are referred to as "consumers." 
* Dataset `dds.discr` 
  * sample of 1,000 DDS consumers (out of a total of ~ 250,000)
  * <span style="color:green">age, gender, race/ethnicity, and DDS annual financial support per consumer</span>
* __Previous research__
  * Researchers examined expenditures on consumers by ethnicity
  * Found that the mean annual expenditures on Hispanics was less than that on White non-Hispanics. 
* Result: an allegation of ethnic discrimination was brought against the California DDS.
* __Question: Are the data sufficient evidence of ethnic discrimination?__

* See Section 1.7.1 for more details


## Load dataset `dds.discr` from package `oibiostat`

* The textbook's datasets are in the R package `oibiostat`
* If you haven't already installed the package `oibiostat`, then first do so using directions in previous slide.

* Load the `oibiostat` package and the dataset `dds.discr`
  * <span style="color:darkorange">the code below needs to be run _every time_ you restart R or knit an Rmd file</span>
```{r}
library(oibiostat)
data("dds.discr")
```

* After loading the dataset `dds.discr` using `data("dds.discr")`, you will see `dds.discr` in the Data list of the Environment window.



## Getting to know the dataset

```{r}
dim(dds.discr)
names(dds.discr)
length(unique(dds.discr$id)) # How many unique id's are there?
```



# Subsetting data

## `filter()` rows that satisfy specified conditions

### `filter()` to select rows

filter data based on rows

- math: `>`,  `<`,  `>=`,  `<=`
-  double = for "is equal to":  `==`
- != (not equal)
-  `&` (and) 
-  `|` (or)


- `is.na()` to filter based on missing values
- `%in%` to filter based on group membership
- `!` in front negates the statement, as in 
    + `!is.na(age)` 
    + `!(ethnicity %in% c("Asian","Black"))`



```{r}
# Note: the output from the command below is not being saved 
# since it's not being assigned to a variable using <-
dds.discr %>% filter(age > 90)   
```


### `filter()` practice

What do these commands do? Try them out:

```{r}
#| eval: false
dds.discr %>% filter(age < 5)
dds.discr %>% filter(age/expenditures < 0.5)    # can do math within filter command
dds.discr %>% filter((age < 15) | (age > 50))

# simultaneously filter on multiple variables
dds.discr %>% filter(age < 20, expenditures > 1000, gender == "Male") 

dds.discr %>% filter(id == 10210) # note the use of == instead of just =
dds.discr %>% filter(gender == "Female")
dds.discr %>% filter(!(age.cohort == "51+"))
dds.discr %>% filter(age.cohort %in% c("0-5", "6-12"))

dds.discr %>% filter(is.na(age))
dds.discr %>% filter(!is.na(age))
```




## Subset by columns

### `select()` to choose columns

- select columns (variables)
- no quotes needed around variable names
- can be used to rearrange columns
- uses special syntax that is flexible and has many options

```{r}
dds.discr %>% select(id, expenditures, ethnicity)
```

### Column selection syntax options

There are many ways to select a set of variable names (columns):

- `var1:var20`: all columns from `var1` to `var20`
- `one_of(c("a", "b", "c"))`: all columns with names in the specified character vector of names

- __Removing columns__
    + `-var1`: remove the column`var1`
    + `-(var1:var20)`: remove all columns from `var1` to `var20`

- __Select by specifying text within column names__
    + `contains("date")`, `contains("_")`: all variable names that contain the specified string of characters
    + `starts_with("a")` or `ends_with("last")`: all variable names that start or end with the specified string

See other examples in the [data transformation cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf).



### `select()` practice

Which columns are selected & in what order using these commands?  
First guess and then try them out.

```{r}
#| eval: false
dds.discr %>% select(id:gender)
dds.discr %>% select(one_of(c("age","expenditures", "notindata")))

dds.discr %>% select(-age.cohort,-gender)
dds.discr %>% select(-(id:gender))

dds.discr %>% select(contains("age"))
dds.discr %>% select(starts_with("a"))
dds.discr %>% select(-contains("a"))
```





# `relocate()` columns to move them around

## `relocate()` to change order of columns

- change the order of columns in dataset
- specified column names get put first, 
  - and unspecified column names after that in original order
- no quotes needed around variable names
- similar options as with `select()`, 
  - plus special ones such as `.before` and `.after`

```{r}
dds.discr %>% relocate(age.cohort, ethnicity)
```



## `relocate()` practice

What order are the columns in using these commands?  
First guess and then try them out.

```{r}
#| eval: false
dds.discr %>% relocate(age:ethnicity)

dds.discr %>% relocate(where(is.numeric))
dds.discr %>% relocate(where(is.factor))
# note: the next command doesn't do anything 
# since there are no character type variables in the dataset
dds.discr %>% relocate(where(is.character))

dds.discr %>% relocate(age,.before = ethnicity)
dds.discr %>% relocate(ethnicity, .after = age.cohort)
dds.discr %>% relocate(age, .after = last_col())
```




# `rename()` columns

- renames column variables

```{r}
dds.discr %>% rename(IDnumber = id)   # order: new_name = old_name
```





# Make new variables

## `mutate()`

Use `mutate()` to add new columns to a tibble
* Many options in how to define new column of data


```{r}
# use = to define new a variable within mutate (not <- or ==) 
newdata <- dds.discr %>% 
  mutate(
    log_expenditures = log(expenditures),
    expend_per_yearage = expenditures / age)   


newdata %>% select(id, age, expenditures, log_expenditures, expend_per_yearage)
```


### `mutate()` practice 

What do the following commands do?  
First guess and then try them out.

```{r}
#| eval: false
dds.discr %>% mutate(age_young = (age < 18))

dds.discr %>% mutate(male = (gender == "Male"))
dds.discr %>% mutate(male = 1 * (gender == "Male"))
```



## `case_when()` to create multi-valued variables

- Example: create age groups based off of the `age` variable
```{r}
dds.discr2 <- dds.discr %>%
  mutate(
    age_group = case_when(
      age < 6 ~ "0 to 5",               # condition ~ new_value
      age >= 6 & age < 13 ~ "6 to 12",
      age >= 13 & age < 18 ~ "13 to 17",
      age >= 18 & age < 22 ~ "18 to 21",
      age >= 22 & age < 51 ~ "22 to 50",
      age >= 51 ~ "51 plus")
    )

dds.discr2 %>% select(age, age_group) %>% head()
dds.discr2 %>% tabyl(age.cohort, age_group)
```




# `arrange()` to order rows

* Use `arrange()` to order the rows by the values in specified columns
  

```{r}
dds.discr %>% arrange(age, expenditures) %>% head(n=3)

dds.discr %>% arrange(desc(age), expenditures) %>% head(n=3)
```




# Reshaping data 


## Wide vs. long data

- __Wide__ data has one row per individual, 
  - with multiple columns for their repeated measurements
- __Long__ data has multiple rows per individual, 
  - with one column for the measurement variable and 
  - another indicating from when/where the repeated measures are from


## DDS example

Mean expenditures by ethnicity and age cohort (from Day 3 slides)


## Example wide toy dataset

Copy and paste the code below into R to create this example dataset

```{r}
SBP_wide <- tibble(id = letters[1:4],
                     sex = c("F", "M", "M", "F"),
                     SBP_v1 = c(130, 120, 130, 119),
                     SBP_v2 = c(110, 116, 136, 106),
                     SBP_v3 = c(112, 122, 138, 118))
SBP_wide
```

- What do you think the data in the table are measures of?
- How can we tell the data are wide?


## Wide to long: `pivot_longer()`

```{r}
SBP_wide
```

For `pivot_longer` we need to __specify__:
- __cols__: which columns to make long
- __names_to__: the name of the variable that will be created from the data stored in the _column names_ 
- __values_to__: the name of the variable that will be created from the data stored in the _cell values_



```{r}
SBP_long <- SBP_wide %>% 
  pivot_longer(
    cols=c(SBP_v1,SBP_v2,SBP_v3), 
    names_to = "visit", 
    values_to = "SBP")
SBP_long
```



## Long to wide: `pivot_wider()`

```{r}
SBP_long
```



For `pivot_wider` we need to __specify__:
- __names_from__: which column contains the names for the *new columns* 
- __values_from__: which column contains the values that will fill in the _cell values_
    
```{r}
SBP_wide2 <- SBP_long %>% 
  pivot_wider(names_from = "visit", 
              values_from = "SBP")
SBP_wide2
```



### Clean up `visit` column in the long data (1/3)


```{r}
SBP_long
```



*Goal*: remove the string "`SBP_`" from the `visit` variable's values.


**Method #1:** tidy the `visit` column *after* making the data long

**Method #2:** tidy the `visit` column *while* making the data long



### Clean up `visit` column in the long data (2/3)


**Method #1:** tidy the `visit` column *after* making the data long

```{r}
SBP_long2 <- SBP_long %>% 
  mutate(
    visit = str_replace(
      visit,
      pattern = "SBP_",
      replacement = "")
    ) 
SBP_long2
```




<!-- Remarks: -->

1. Note that `mutate` is replacing the existing `visit` column with new values
1. If I wanted to keep the original `visit` column instead of overwriting it, I would call the new column something else, such as `visit_clean`
1. Within `str_replace()`, double quotes need to be used around the characters specifying 1) the string of text to replace (`"SBP_"`) and 2) what to replace the text with, where `""` is used for *no text*
1. Could instead use `str_remove()`:
`mutate(visit = str_remove(visit,"SBP_"))` 







### Clean up `visit` column in the long data (3/3)


**Method #2:** tidy the `visit` column *while* making the data long

```{r}
SBP_long3 <- SBP_wide %>% 
 pivot_longer(cols = c(SBP_v1, SBP_v2,SBP_v3), 
              names_to = "visit", 
              names_prefix = "SBP_", #<<
              values_to = "SBP")
SBP_long3
```




Remarks:

1. Note the new parameter `names_prefix` specifying what the prefix is that *needs to be stripped*.
1. More complex `pivot_longer()` examples are shown at
https://tidyr.tidyverse.org/articles/pivot.html  
*(the end of the url is articles/pivot.html)*






## Specifying `cols` in `pivot_longer`


In the example creating `SBP_long`, the columns to make the tibble longer by were explicitly listed using `cols=c(SBP_v1,SBP_v2,SBP_v3)`.

```{r}
SBP_long <- SBP_wide %>% 
 pivot_longer(cols = c(SBP_v1, SBP_v2,SBP_v3), #<<
              names_to = "visit", 
              values_to = "SBP")
SBP_long
```






However, we can specify the columns in many different ways, just like with `select()`:

* `cols = c(SBP_v1:SBP_v3)`
* `cols = c(-id, -sex)`
* `cols = starts_wth("SBP")`
* `cols = contains("SBP")`





## Notes on `pivot_*()` commands

* `pivot_longer()` and `pivot_wider()` are relatively new commands
* previously we used `gather()` and `spread()`, 
  - which have different function parameters and are less intuitive to use
* if you search for help making data longer or wider, 
  - you might still see references for `gather()` and `spread()`
* see my [workshop slides](https://jminnier-berd-r-courses.netlify.com/02-data-wrangling-tidyverse/02_data_wrangling_slides_part2.html#26) for `gather()` and `spread()` usage


* see https://tidyr.tidyverse.org/articles/pivot.html for more `pivot` examples





# Summary of data wrangling commands

command | purpose
--|-
 `%>%` | join (pipe) together commands
`filter()` | subset rows
`arrange()` | sort rows
`select()` | select or rearrange columns
`rename()` | columns
`mutate()` | create new columns
`tabyl()` | summarize categorical data
`summarize()` | summarize data; for both categorical and numerical data
`group_by()` | group data by a categorical variable
`pivot_longer` | make a wide dataset long; 
`pivot_wider` | to make a long dataset wide

<!-- `bind_rows` | merge rows from different datasets -->
<!-- `left_join()` | merge different datasets; also `full_join()`, `inner_join()`, etc. -->


