---
title: "Day 9: Confidence intervals (4.2)"
subtitle: "BSTA 511/611"
pagetitle: "Day 9 BSTA 511/611"
author: "Meike Niederhausen, PhD"
institute: "OHSU-PSU School of Public Health"
date: "10/29/2025"
categories: ["Week 5"]
format: 
  html:
    link-external-newwindow: true
    toc: true
execute:
  echo: true
# editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo = TRUE, fig.height=3, fig.width=5, message = F)
```

## Load packages

* Packages need to be loaded _every time_ you restart R or render an Qmd file

```{r}
# run these every time you open Rstudio
library(tidyverse)    
library(oibiostat)
library(ggridges)
library(janitor)
library(rstatix)
library(knitr)
library(gtsummary)
library(moderndive) # NEW!!

set.seed(456)
```


- You can check whether a package has been loaded or not 
  - by looking at the Packages tab and 
  - seeing whether it has been checked off or not
  
  
# Last time -> Goals for today

## Day 8: Section 4.1
* Sampling from a population
  * population __parameters__ vs. __point estimates__
  * sampling variation

<br>

* __Sampling distribution__ of a mean
* __Central Limit Theorem__

## Day 9: Section 4.2

What are __Confidence Intervals__?

* How to __calculate__ CI's?
* How to __interpret__ & __NOT__ interpret CI's?

<br>

* What if we don't know $\sigma$?
* Student's __t-distribution__


# Our hypothetical population: YRBSS

__Youth Risk Behavior Surveillance System (YRBSS)__

* Yearly survey conducted by the US Centers for Disease Control (CDC) 
* Measures health-related activity in high-school aged youth
* `yrbss` contains responses from _n_ = 13,583 participants in 2013 for a subset of the variables included in the complete survey data

```{r}
library(oibiostat)
data("yrbss")  #load the data
# ?yrbss
```

```{r}
dim(yrbss)
```

```{r}
names(yrbss)
```


## Transform height & weight from metric to to standard

Also, drop missing values and add a column of id values
```{r}
yrbss2 <- yrbss %>%                 # save new dataset with new name
  mutate(                           # add variables for 
    height.ft = 3.28084*height,     #     height in feet
    weight.lb = 2.20462*weight      #     weight in pounds
  ) %>% 
  drop_na(height.ft, weight.lb) %>% # drop rows w/ missing height/weight values
  mutate(id = 1:nrow(.)) %>%        # add id column
  select(id, height.ft, weight.lb)  # restrict dataset to columns of interest

head(yrbss2)  
dim(yrbss2)
# number of rows deleted that had missing values for height and/or weight:
nrow(yrbss) - nrow(yrbss2) 
```

## `yrbss2`: stats for height in feet 

```{r}
summary(yrbss2)
(mean_height.ft <- mean(yrbss2$height.ft))
(sd_height.ft <- sd(yrbss2$height.ft))
```

## Take 10,000 samples of size n = 30 from `yrbss2`

```{r}
samp_n30_rep10000 <- yrbss2 %>%
  rep_sample_n(size = 30, 
               reps = 10000,
               replace = FALSE)
samp_n30_rep10000
```

```{r}
means_hght_samp_n30_rep10000 <- 
  samp_n30_rep10000 %>% 
  group_by(replicate) %>% 
  summarise(mean_height = 
            mean(height.ft))

means_hght_samp_n30_rep10000
```


## Simulated sampling distribution for n = 30 using 10,000 sample mean heights

```{r}
ggplot(
  means_hght_samp_n30_rep10000, 
  aes(x = mean_height)) + 
  geom_histogram()
```

CLT tells us that we can model the sampling distribution of mean heights using a normal distribution.

```{r}
#| fig.height: 3
#| fig.width: 6
mu <- 5.55
SE <- 0.34/sqrt(30)
sig <- round(SE, 2)

ggplot(data.frame(x = c(mu-4*sig, mu+4*sig)), aes(x = x)) + 
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = sig)) + 
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = sig), 
                xlim = c(mu-4*sig, mu-2*sig), 
                geom = "area", fill = "darkblue") +
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = sig), 
                xlim = c(mu+2*sig, mu+4*sig), 
                geom = "area", fill = "darkblue") +
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = sig), 
                xlim = c(mu-2*sig, mu-1*sig), 
                geom = "area", fill = "darkgreen") +
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = sig), 
                xlim = c(mu+1*sig, mu+2*sig), 
                geom = "area", fill = "darkgreen")+
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = sig), 
                xlim = c(mu-1*sig, mu+1*sig), 
                geom = "area", fill = "orange") +
  scale_x_continuous(name ="mean height (ft)", 
                    breaks=c(mu-2*sig,mu-1*sig,mu, mu+1*sig, mu+2*sig))  +
  labs(title = "Sampling distribution", y = "") +
  scale_y_continuous(labels = NULL, breaks = NULL)
```


# Confidence interval (CI) for the mean $\mu$

$$\overline{x}\ \pm\ z^*\times \text{SE}$$

where 

* $SE = \frac{\sigma}{\sqrt{n}}$
* $z^*$ depends on the confidence level
  * For a 95% CI, $z^*$ is chosen such that 95% of the standard normal curve is between $-z^*$ and $z^*$

```{r}
qnorm(.975)
qnorm(.995)
```

[When can this be applied?]{style="color:green"}


## Example: CI for mean height

* A random sample of 30 high schoolers has mean height 5.6 ft. 
* Find the 95% confidence interval for the population mean, assuming that the population standard deviation is 0.34 ft. 





## How to interpret a CI?

Simulating Confidence Intervals: http://www.rossmanchance.com/applets/ConfSim.html

[Actual interpretation:]{style="color:darkorange"}

* If we were to 
    * __repeatedly take random samples__ from a population and
    * calculate a 95% CI for each random sample,
* then we would __expect 95% of our CI's to contain the true population parameter__ $\mu$.

[What we typically write as "shorthand":]{style="color:darkorange"}

* We are 95% _confident_  that (the 95% confidence interval) captures the value of the population parameter.

[WRONG interpretation:]{style="color:red"}

* There is a 95% _chance_  that (the 95% confidence interval) captures the value of the population parameter.
    * For one CI on its own, it either does or doesn't contain the population parameter with probability 0 or 1. We just don't know which!


## Interpretation of our heights CI

Correct interpretation:

* We are 95% _confident_ that the mean height for high schoolers is between 5.43 and 5.67 feet.

[*WRONG:*]{style="color:red"}

* There is a 95% _chance_ that the mean height for high schoolers is between 5.43 and 5.67 feet.


# What if we don't know $\sigma$ ? 

* In real life, we don't know what the population sd is ( $\sigma$ )

* If we replace $\sigma$ with $s$ in the SE formula, we add in additional variability to the SE!
$$\frac{\sigma}{\sqrt{n}} ~~~~\textrm{vs.} ~~~~ \frac{s}{\sqrt{n}}$$

* Thus when using $s$ instead of $\sigma$ when calculating the SE, we __need a different probability distribution__ with thicker tails than the normal distribution.

    * [In practice this will mean using a different value than 1.96 when calculating the CI.]{style="color:green"}


## Student’s t-distribution

The __Student’s t-distribution__:

* Is bell shaped and symmetric with mean = 0.
* Its tails are a thicker than that of a normal distribution
    * The "thickness" depends on its __degrees of freedom__: $df = n–1$ , where n = sample size.
* As the degrees of freedom (sample size) increase, 
    * the tails are less thick, and 
    * the t-distribution is more like a normal distribution
    * in theory, with an infinite sample size the _t_-distribution is a normal distribution.


## Calculating the CI for the population mean using $s$

CI for $\mu$:

$$\bar{x} \pm t^*\cdot\frac{s}{\sqrt{n}}$$

where $t^*$ is determined by the t-distribution and dependent on the  
__df =__ $n-1$ and the confidence level

* `qt` gives the quartiles for a t-distribution. Need to specify 
    * the percent under the curve to the left of the quartile 
    * the degrees of freedom = n-1

* Note in the R output to the right that $t^*$ gets closer to 1.96 as the sample size increases.

```{r}
qt(.975, df=9)  # df = n-1
qt(.975, df=49)
qt(.975, df=99)
qt(.975, df=999)
```



## Example: CI for mean height (revisited)

* A random sample of 30 high schoolers has mean height 5.6 ft and standard deviation 0.34 ft. 
* Find the 95% confidence interval for the population mean. 

# $z$ vs $t$?? <br> ([& important comment about Chapter 4 of textbook]{style="color:green"})


## Textbook's rule of thumb

* (Ch 4) If $n \geq 30$ and population distribution not strongly skewed:
    * __Use normal distribution__
    * __No matter if using $\sigma$ or $s$ for the $SE$__
    * If there is skew or some large outliers, then need  $n \geq 50$
* (Ch 5) If $n < 30$ and data approximately symmetric with no large outliers:
    * Use Student's t-distribution

## BSTA 511 rule of thumb

* Use [normal distribution ONLY if know $\sigma$]{style="color:darkorange"}
    * If [using $s$ for the $SE$, then use the Student's t-distribution]{style="color:purple"}

For either case, can apply if either

* $n \geq 30$ and population distribution not strongly skewed
    * If there is skew or some large outliers, then $n \geq 50$ gives better estimates
* $n < 30$ and data approximately symmetric with no large outliers

_If do not know population distribution, then check the distribution of the data._

