---
title: "Day 12: Inference for a single proportion or difference of two (independent) proportions (Sections 8.1-8.2)"
subtitle: "BSTA 511/611"
pagetitle: "Day 12 BSTA 511/611"
author: "Meike Niederhausen, PhD"
institute: "OHSU-PSU School of Public Health"
date: "11/12/2025"
categories: ["Week 7"]
format: 
  html:
    link-external-newwindow: true
    toc: true
    code-fold: show
    code-tools: true
    source: repo
execute:
  echo: true
  freeze: auto  # re-render only when source changes
# editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo = TRUE, fig.height=3, fig.width=5, message = F)
```

## Load packages

* Packages need to be loaded _every time_ you restart R or render an Qmd file

```{r}
# run these every time you open Rstudio
library(tidyverse)    
library(oibiostat)
library(janitor)
library(rstatix)
library(knitr)
library(gtsummary)
library(moderndive)
library(gt)
library(broom) 
library(here) 
library(pwr) # new-ish

```


- You can check whether a package has been loaded or not 
  - by looking at the Packages tab and 
  - seeing whether it has been checked off or not


## MoRitz's tip of the day: [code folding]{style="color:darkorange"}

* With code folding we can hide or show the code in the html output by clicking on the `Code` buttons in the html file.
* Note the `</> Code` button on the top right of the html output.


* See the new options in the yaml above (in the .qmd file).

>code-fold: show
 code-tools: true
 source: repo

See more information at <https://quarto.org/docs/output-formats/html-code.html#folding-code>



## Where are we?

CI's and hypothesis tests for different scenarios:

$$\text{point estimate} \pm z^*(or~t^*)\cdot SE,~~\text{test stat} = \frac{\text{point estimate}-\text{null value}}{SE}$$

Day | Book | Population <br> parameter | Symbol | Point estimate | Symbol | SE
--|--|--|--|--|--|--
10 | 5.1 | Pop mean | $\mu$ | Sample mean | $\bar{x}$ | $\frac{s}{\sqrt{n}}$
10 | 5.2 | Pop mean of paired diff | $\mu_d$ or $\delta$ | Sample mean of paired diff | $\bar{x}_{d}$  | **$\frac{s_d}{\sqrt{n}}$**
11 | 5.3 | [Diff in pop <br> means]{style="color:black"} | [$\mu_1-\mu_2$]{style="color:black"} | [Diff in sample <br> means]{style="color:black"} | [$\bar{x}_1 - \bar{x}_2$]{style="color:black"}  | [**$\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$ or pooled**]{style="color:black"}
12 | 8.1 | [Pop proportion]{style="color:green"} | [$p$]{style="color:green"} | [Sample prop]{style="color:green"} | [$\widehat{p}$]{style="color:green"}  | [**???**]{style="color:red"}
12 | 8.2 | [Diff in pop <br> proportions]{style="color:green"} | [$p_1-p_2$]{style="color:green"} | [Diff in sample <br> proportions]{style="color:green"} | [$\widehat{p}_1-\widehat{p}_2$]{style="color:green"} | [**???**]{style="color:red"}




## Goals for today (Sections 8.1-8.2)

* Statistical inference for a single proportion or the difference of two (independent) proportions
    1. Sampling distribution for a proportion or difference in proportions
    
    1. What are $H_0$ and $H_a$?
    
    1. What are the SE's for $\hat{p}$ and $\hat{p}_1-\hat{p}_2$?
    
    1. Hypothesis test
    
    1. Confidence Interval
    
    1. How are the SE's different for a hypothesis test & CI?
    
    1. How to run proportions tests in R
    
    1. Power & sample size for proportions tests (extra material)


# Motivating example

__One proportion__

* A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year. 
    * What is the CI for the proportion?
    * The study also reported that 36% of noncollege young males had participated in sports betting. Is the proportion for male college students different from 0.36?

__Two proportions__

* There were 214 men in the sample of noncollege young males (36% participated in sports betting in the previous year).
* Compare the difference in proportions between the college and noncollege young males.
    * CI & Hypothesis test


Barnes GM, Welte JW, Hoffman JH, Tidwell MC. [Comparisons of gambling and alcohol use among college students and noncollege young people in the United States](https://www.tandfonline.com/doi/full/10.1080/07448480903540499?journalCode=vach20). J Am Coll Health. 2010 Mar-Apr;58(5):443-52. doi: 10.1080/07448480903540499. PMID: 20304756; PMCID: PMC4104810.




## Steps in a Hypothesis Test

1. Set the __[level of significance]{style="color:darkorange"}__ $\alpha$

1. Specify the __[null]{style="color:darkorange"}__ ( $H_0$ ) and __[alternative]{style="color:darkorange"}__ ( $H_A$ ) __[hypotheses]{style="color:darkorange"}__
    1. In symbols
    1. In words
    1. Alternative: one- or two-sided?

1. Calculate the __[test statistic]{style="color:darkorange"}__. 

1. Calculate the __[p-value]{style="color:darkorange"}__ based on the observed test statistic and its sampling distribution

1. Write a __[conclusion]{style="color:darkorange"}__ to the hypothesis test
    1. Do we reject or fail to reject $H_0$?
    1. Write a conclusion in the context of the problem


# Step 2: Null & Alternative Hypotheses 

Null and alternative hypotheses in __words__ and in __symbols__.


__One sample test__

* $H_0$: The population proportion of young male college students that participated in sports betting in the previous year is 0.36.

* $H_A$: The population proportion of young male college students that participated in sports betting in the previous year is not 0.36.

\begin{align}
H_0:& p = 0.36\\
H_A:& p \neq 0.36\\
\end{align}


__Two samples test__

* $H_0$: The difference in population proportions of young male college and noncollege students that participated in sports betting in the previous year is 0.

* $H_A$: The difference in population proportions of young male college and noncollege students that participated in sports betting in the previous year is not 0.

\begin{align}
H_0:& p_{coll} - p_{noncoll} = 0\\
H_A:& p_{coll} - p_{noncoll} \neq 0\\
\end{align}


# One proportion inference

## Sampling distribution of $\hat{p}$ 

* $\hat{p}=\frac{X}{n}$ where $X$ is the number of "successes" and $n$ is the sample size.
* $X \sim Bin(n,p)$, where $p$ is the population proportion.
* For $n$ "big enough", the normal distribution can be used to approximate a binomial distribution:

$$Bin(n,p) \rightarrow N\Big(\mu = np, \sigma = \sqrt{np(1-p)} \Big)$$

* Since $\hat{p}=\frac{X}{n}$ is a linear transformation of $X$, we have for large n: 

$$\hat{p} \sim N\Big(\mu_{\hat{p}} = p, \sigma_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}} \Big)$$

* [*How we apply this result to CI's and test statistics is different!!!*]{style="color:purple"}



## Step 3: Test statistic 

Sampling distribution of $\hat{p}$ if we assume $H_0: p=p_0$ is true:

$$\hat{p} \sim N\Big(\mu_{\hat{p}} = p, \sigma_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}} \Big) 
\sim N\Big(
\mu_{\hat{p}}=p_0, \sigma_{\hat{p}}=\sqrt{\frac{p_0\cdot(1-p_0)}{n}}
\Big)$$


Test statistic for a one sample proportion test:

$$
\text{test stat} = \frac{\text{point estimate}-\text{null value}}{SE}
= z_{\hat{p}} = \frac{\hat{p} - p_0}{\sqrt{\frac{p_0\cdot(1-p_0)}{n}}}
$$

<hr>

__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year.   
<br>
What is the test statistic when testing $H_0: p=0.36$ vs. 
$H_A: p \neq 0.36$? 


```{r}
p0 <- 0.36
n <- 269
n*.35
(ph <- 94/n)

(SEp <- sqrt(p0*(1-p0)/n))
(zp <- (ph-p0)/SEp)
```

\begin{align}
z_{\hat{p}} &= \frac{94/269 - 0.36}{\sqrt{\frac{0.36\cdot(1-0.36)}{269}}} \\
& `r zp`
\end{align}



## Step "3b": Conditions satisfied?

__Conditions__:

1. _Independent observations_ 
    * The observations were collected independently.

1. The number of __expected successes and expected failures is at least 10__.
    * $n_1 p_0 \ge 10, \ \ n_1(1-p_0)\ge 10$

<hr>

__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year.  
<br>
Testing $H_0: p=0.36$ vs. $H_A: p \neq 0.36$.  
<br>
Are the conditions satisfied?



## Step 4: p-value

The __[p-value]{style="color:darkorange"}__ is the __probability__ of obtaining a test statistic _just as extreme or more extreme_ than the observed test statistic assuming the null hypothesis $H_0$ is true. 


```{r}
#| fig.width: 4
#| fig.height: 2.5
#| echo: false
# specify upper and lower bounds of shaded region below
mu <- 0.36
std <- 0.03

# The following figure is only an approximation of the 
# sampling distribution since I used a normal instead
# of t-distribution to make it.

ggplot(data.frame(x = c(mu-5*std, mu+5*std)), aes(x = x)) + 
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = std)) + 
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks=c(mu, mu - 0.03*(1:5), mu + 0.03*(1:5))) +
  theme(axis.text.x=element_text(angle = -30, hjust = 0)) +
  labs(y = "", 
       x = "p-hat distribution") +
  geom_vline(xintercept = c(0.35, 0.37), 
             color = "red")
```

```{r}
#| fig.height: 2.5
#| fig.width: 4
#| echo: false
ggplot(data = data.frame(x = c(-5, 5)), aes(x)) + 
  stat_function(fun = dnorm, 
                args = list(mean = 0, sd = 1)) + 
  ylab("") + 
  xlab("z-dist") +
  scale_y_continuous(breaks = NULL) + 
  scale_x_continuous(breaks=c(mu, mu - (1:5), mu + (1:5))) +
  geom_vline(xintercept = c(-0.34, 0.34), 
             color = "red")
```


Calculate the _p_-value:

\begin{align}
2 &\cdot P(\hat{p}<0.35) \\
&= 2 \cdot P\Big(Z_{\hat{p}} < \frac{94/269 - 0.36}{\sqrt{\frac{0.36\cdot(1-0.36)}{269}}}\Big)\\
&=2 \cdot P(Z_{\hat{p}} < -0.3607455)\\
&= `r 2*pnorm(-0.3607455)`
\end{align}

```{r}
2*pnorm(-0.3607455)
```



## Step 5: Conclusion to hypothesis test

\begin{align}
H_0:& p = 0.36\\
H_A:& p \neq 0.36\\
\end{align}

* Recall the $p$-value = 0.7182897
* Use $\alpha$ = 0.05.
* Do we reject or fail to reject $H_0$?

__Conclusion statement__:

* Stats class conclusion
    * There is insufficient evidence that the (population) proportion of young male college students that participated in sports betting in the previous year is different than 0.36 ( $p$-value = 0.72).

* More realistic manuscript conclusion: 
    * In a sample of 269 male college students, 35% had participated in sports betting in the previous year, which is not different from 36% ( $p$-value = 0.72).



## 95% CI for population proportion

What to use for SE in CI formula?

$$\hat{p} \pm z^* \cdot SE_{\hat{p}}$$

Sampling distribution of $\hat{p}$:

$$\hat{p} \sim N\Big(\mu_{\hat{p}} = p, \sigma_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}} \Big)$$

Problem: We don't know what $p$ is - it's what we're estimating with the CI.  
Solution: [approximate $p$ with $\hat{p}$]{style="color:purple"}:

$$SE_{\hat{p}} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

<hr>

__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year.   
Find the 95% CI for the population proportion. 

\begin{align}
94/269 &\pm 1.96 \cdot SE_{\hat{p}}\\
SE_{\hat{p}} &= \sqrt{\frac{(94/269)(1-94/269)}{269}}
\end{align}

__Interpretation__:  
We are 95% confident that the (population) proportion of young male college students that participated in sports betting in the previous year is in (0.29, 0.41).


## Conditions for one proportion: test vs. CI


[__Hypothesis test conditions__]{style="color:green"}

1. _Independent observations_ 
    * The observations were collected independently.

1. The number of __expected__ successes and __expected__ failures is at least 10.

$$n_1 p_0 \ge 10, \ \ n_1(1-p_0)\ge 10$$



[__Confidence interval conditions__]{style="color:purple"}

1. _Independent observations_ 
    * The observations were collected independently.

1. The number of successes and failures is at least 10:

$$n_1\hat{p}_1 \ge 10, \ \ n_1(1-\hat{p}_1)\ge 10$$


# Inference for difference of two independent proportions $\hat{p}_1-\hat{p}_2$


## Sampling distribution of $\hat{p}_1-\hat{p}_2$ 

* $\hat{p}_1=\frac{X_1}{n_1}$ and $\hat{p}_2=\frac{X_2}{n_2}$, 
    * $X_1$ & $X_2$ are the number of "successes"
    * $n_1$ & $n_2$ are the sample sizes of the 1st & 2nd samples 

<br>

* Each $\hat{p}$ can be approximated by a normal distribution, for "big enough" $n$
* Since the difference of independent normal random variables is also normal, it follows that for "big enough" $n_1$ and $n_2$ 

$$\hat{p}_1 - \hat{p}_2 \sim N \Big(\mu_{\hat{p}_1 - \hat{p}_2} = p_1 - p_2, ~~
\sigma_{\hat{p}_1 - \hat{p}_2} =
\sqrt{
\frac{p_1\cdot(1-p_1)}{n_1} + \frac{p_2\cdot(1-p_2)}{n_2}} 
\Big)$$

where $p_1$ & $p_2$ are the population proportions, respectively.

* [*How we apply this result to CI's and test statistics is different!!!*]{style="color:purple"}



## Step 3: Test statistic (1/2)

Sampling distribution of $\hat{p}_1 - \hat{p}_2$:
$$\hat{p}_1 - \hat{p}_2 \sim N \Big(\mu_{\hat{p}_1 - \hat{p}_2} = p_1 - p_2, ~~
\sigma_{\hat{p}_1 - \hat{p}_2} =
\sqrt{
\frac{p_1\cdot(1-p_1)}{n_1} + \frac{p_2\cdot(1-p_2)}{n_2}} 
\Big)$$

Since we assume $H_0: p_1 - p_2 = 0$ is true, we "pool" the proportions of the two samples to calculate the SE:

$$\text{pooled proportion} = \hat{p}_{pool} = \dfrac{\text{total number of successes} }{ \text{total number of cases}} = \frac{x_1+x_2}{n_1+n_2}$$ 


Test statistic:

$$
\text{test statistic} = z_{\hat{p}_1 - \hat{p}_2} = \frac{\hat{p}_1 - \hat{p}_2 - 0}{\sqrt{\frac{\hat{p}_{pool}\cdot(1-\hat{p}_{pool})}{n_1} + \frac{\hat{p}_{pool}\cdot(1-\hat{p}_{pool})}{n_2}}}
$$


## Step 3: Test statistic (2/2)


$$
\text{test statistic} = z_{\hat{p}_1 - \hat{p}_2} = \frac{\hat{p}_1 - \hat{p}_2 - 0}{\sqrt{\frac{\hat{p}_{pool}\cdot(1-\hat{p}_{pool})}{n_1} + \frac{\hat{p}_{pool}\cdot(1-\hat{p}_{pool})}{n_2}}}
$$

$$\text{pooled proportion} = \hat{p}_{pool} = \dfrac{\text{total number of successes} }{ \text{total number of cases}} = \frac{x_1+x_2}{n_1+n_2}$$ 

<hr>

__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year, and out of 214 noncollege young males 36% had.  
What is the test statistic when testing $H_0: p_{coll} - p_{noncoll} = 0$ vs. 
$H_A: p_{coll} - p_{noncoll} \neq 0$? 

\begin{align}
z_{\hat{p}_1 - \hat{p}_2} &= \frac{94/269 - 77/214-0}{\sqrt{0.354\cdot(1-0.354)(\frac{1}{269}+\frac{1}{214})}}\\
&=-0.2367497
\end{align}


## Step "3b": Conditions satisfied?


__Conditions__:

* _Independent observations & samples_
    * The observations were collected independently. 
    * In particular, observations from the two groups weren't paired in any meaningful way.

* The number of expected successes and expected failures is at least 10 _for each group_ - using the pooled proportion:
    * $n_1\hat{p}_{pool} \ge 10, \ \ n_1(1-\hat{p}_{pool}) \ge 10$
    * $n_2\hat{p}_{pool} \ge 10, \ \ n_2(1-\hat{p}_{pool}) \ge 10$



__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year, and out of 214 noncollege young males 36% had.   
Testing $H_0: p_{coll} - p_{noncoll} = 0$ vs. 
$H_A: p_{coll} - p_{noncoll} \neq 0$? .  
Are the conditions satisfied?


## Step 4: p-value

The __[p-value]{style="color:darkorange"}__ is the __probability__ of obtaining a test statistic _just as extreme or more extreme_ than the observed test statistic assuming the null hypothesis $H_0$ is true. 


```{r}
#| fig.width: 4
#| fig.height: 2.5
#| echo: false
# specify upper and lower bounds of shaded region below
mu <- 0
std <- 0.02

# The following figure is only an approximation of the 
# sampling distribution since I used a normal instead
# of t-distribution to make it.

ggplot(data.frame(x = c(mu-5*std, mu+5*std)), aes(x = x)) + 
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = std)) + 
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks=c(mu, mu - 0.02*(1:5), mu + 0.02*(1:5))) +
  theme(axis.text.x=element_text(angle = -30, hjust = 0)) +
  labs(y = "", 
       x = "p-hat distribution") +
  geom_vline(xintercept = c(-.01, 0.01), 
             color = "red")
```

```{r}
#| fig.height: 2.5
#| fig.width: 4
#| echo: false
ggplot(data = data.frame(x = c(-5, 5)), aes(x)) + 
  stat_function(fun = dnorm, 
                args = list(mean = 0, sd = 1)) + 
  ylab("") + 
  xlab("z-dist") +
  scale_y_continuous(breaks = NULL) + 
  scale_x_continuous(breaks=c(mu, mu - (1:5), mu + (1:5))) +
  geom_vline(xintercept = c(-0.45, 0.45), 
             color = "red")
```


Calculate the _p_-value:

\begin{align}
2 &\cdot P(\hat{p}_1 - \hat{p}_2<0.35-0.36) \\
&= 2 \cdot P\Big(Z_{\hat{p}_1 - \hat{p}_2} < \\
&\frac{94/269 - 77/214-0}{\sqrt{0.354\cdot(1-0.354)(\frac{1}{269}+\frac{1}{214})}}\Big)\\
&=2 \cdot P(Z_{\hat{p}} < -0.2367497)
\end{align}
```{r}
2*pnorm(-0.2367497)
```




## Step 5: Conclusion to hypothesis test

\begin{align}
H_0:& p_{coll} - p_{noncoll} = 0\\
H_A:& p_{coll} - p_{noncoll} \neq 0\\
\end{align}

* Recall the $p$-value = 0.812851
* Use $\alpha$ = 0.05.
* Do we reject or fail to reject $H_0$?

__Conclusion statement__:

* Stats class conclusion
    * There is insufficient evidence that the difference in (population) proportions of young male college and noncollege students that participated in sports betting in the previous year are different ( $p$-value = 0.81).

* More realistic manuscript conclusion: 
    * 35% of young male college students (n=269) and 36% of noncollege young males (n=214) participated in sports betting in the previous year ( $p$-value = 0.81).



## 95% CI for population difference in proportions

What to use for SE in CI formula?

$$\hat{p}_1 - \hat{p}_2 \pm z^* \cdot SE_{\hat{p}_1 - \hat{p}_2}$$

SE in sampling distribution of $\hat{p}_1 - \hat{p}_2$

$$\sigma_{\hat{p}_1 - \hat{p}_2} =
\sqrt{
\frac{p_1\cdot(1-p_1)}{n_1} + \frac{p_2\cdot(1-p_2)}{n_2}} $$

Problem: We don't know what $p$ is - it's what we're estimating with the CI.  
Solution: [approximate $p_1$, $p_2$ with $\hat{p}_1$, $\hat{p}_2$]{style="color:purple"}:

$$SE_{\hat{p}_1 - \hat{p}_2} = \sqrt{
\frac{\hat{p}_1\cdot(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2\cdot(1-\hat{p}_2)}{n_2}}$$

<hr>

__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year, and out of 214 noncollege young males 36% had.
Find the 95% CI for the difference in population proportions. 

$$\frac{94}{269} - \frac{77}{214} \pm 1.96 \cdot SE_{\hat{p}_1 - \hat{p}_2}$$

\begin{align}
& SE_{\hat{p}_1 - \hat{p}_2}=\\
& \sqrt{
\frac{94/269 \cdot (1-94/269)}{269} + 
\frac{77/214 \cdot (1-77/214)}{214}}
\end{align}

__Interpretation__:  
We are 95% confident that the difference in (population) proportions of young male college and noncollege students that participated in sports betting in the previous year is in (-0.127, 0.106).



## Conditions for difference in proportions: test vs. CI


[__Hypothesis test conditions__]{style="color:green"}

1. _Independent observations & samples_
    * The observations were collected independently. 
    * In particular, observations from the two groups weren't paired in any meaningful way.


2. The number of __expected__ successes and __expected__ failures is at least 10 _for each group_ - using the pooled proportion:
    * $n_1\hat{p}_{pool} \ge 10, \ \ n_1(1-\hat{p}_{pool}) \ge 10$
    * $n_2\hat{p}_{pool} \ge 10, \ \ n_2(1-\hat{p}_{pool}) \ge 10$

[__Confidence interval conditions__]{style="color:purple"}

1. _Independent observations & samples_
    * The observations were collected independently. 
    * In particular, observations from the two groups weren't paired in any meaningful way.

2. The number of successes and failures is at least 10 _for each group_.
    * $n_1\hat{p}_1 \ge 10, \ \ n_1(1-\hat{p}_1) \ge 10$
    * $n_2\hat{p}_2 \ge 10, \ \ n_2(1-\hat{p}_2) \ge 10$



# R: 1- and 2-sample proportions tests

`prop.test(x, n, p = NULL,`  
&nbsp; &nbsp; &nbsp;     `alternative = c("two.sided", "less", "greater"),`   
&nbsp; &nbsp; &nbsp;     `conf.level = 0.95,`  
&nbsp; &nbsp; &nbsp;     `correct = TRUE)`

* 2 options for data input 
    1. Summary counts
        * `x` = vector with counts of "successes" 
        * `n` = vector with sample size in each group
    2. Dataset
        * `x = table()` of dataset
        * Need to create a dataset based on the summary stats if do not already have one
* Continuity correction

# R: 1-sample proportion test

"1-prop z-test"


## Summary stats input for 1-sample proportion test

__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year.   
Test $H_0: p=0.36$ vs. $H_A: p \neq 0.36$? 

```{r}
.35*269 # number of "successes"; round this value

prop.test(x = 94, n = 269,              # x = # successes & n = sample size
          p = 0.36,                     # null value p0
          alternative = "two.sided",    # 2-sided alternative
          correct = FALSE)              # no continuity correction
```

Can `tidy()` test output:
```{r}
prop.test(x = 94, n = 269, p = 0.36, alternative = "two.sided", correct = FALSE) %>% 
  tidy() %>% gt()
```


## Dataset input for 1-sample proportion test (1/2)

Since we don't have a dataset, we first need to create a dataset based on the results:

>"out of 269 male college students, 35% had participated in sports betting in the previous year"

```{r}
.35*269 # number of "successes"; round this value

SportsBet1 <- tibble(
  Coll = c(rep("Bet", 94), 
           rep("NotBet",269-94))
  )

glimpse(SportsBet1)

SportsBet1 %>% tabyl(Coll)
```


R code for proportions test requires input as a base R `table`:
```{r}
table(SportsBet1$Coll)
```


## Dataset input for 1-sample proportion test (2/2)

* When using a dataset, `prop.test` requires the input `x` to be a table
* Note that we do not also specify `n` since the table already includes all needed information.


```{r}
prop.test(x = table(SportsBet1$Coll),   # table() of data
          p = 0.36,                     # null value p0
          alternative = "two.sided",    # 2-sided alternative
          correct = FALSE)              # no continuity correction


```

Compare output with summary stats method:

```{r}
prop.test(x = 94, n = 269,              # x = # successes & n = sample size
          p = 0.36,                     # null value p0
          alternative = "two.sided",    # 2-sided alternative
          correct = FALSE)  %>%         # no continuity correction
  tidy() %>% gt()
```


## Continuity correction: 1-prop z-test with vs. without CC


* Recall that when we approximated the 
* binomial distribution with a normal distribution to calculate a probability, 
* that we included a continuity correction (CC) 
* to account for approximating a discrete distribution with a continuous distribution.



```{r}
prop.test(x = table(SportsBet1$Coll), 
          alternative = "two.sided",
          p = 0.36, 
          correct = FALSE) %>% 
  tidy() %>% gt()
```


```{r}
prop.test(x = table(SportsBet1$Coll), 
          alternative = "two.sided",
          p = 0.36, 
          correct = TRUE) %>% 
  tidy() %>% gt()
```


Differences are small when sample sizes are large.



# R: 2-samples proportion test

"2-prop z-test"

## Summary stats input for 2-samples proportion test

__Example:__ A 2010 study found that out of 269 male college students, 35% had participated in sports betting in the previous year, and out of 214 noncollege young males 36% had. Test $H_0: p_{coll} - p_{noncoll} = 0$ vs. $H_A: p_{coll} - p_{noncoll} \neq 0$.  

```{r}
# round the number of successes:
.35*269 # number of "successes" in college students
.36*214 # number of "successes" in noncollege students

NmbrBet <- c(94, 77)                    # vector for # of successes in each group
TotalNmbr <- c(269, 214)                # vector for sample size in each group

prop.test(x = NmbrBet,                  # x is # of successes in each group 
          n = TotalNmbr,                # n is sample size in each group
          alternative = "two.sided",    # 2-sided alternative
          correct = FALSE)              # no continuity correction
```


## Dataset input for 2-samples proportion test (1/2)

Since we don't have a dataset, we first need to create a dataset based on the results:

>"out of 269 male college students, 35% had participated in sports betting in the previous year, and out of 214 noncollege young males 36% had"


```{r}
.35*269 # number of "successes"
.36*214 # round these value

SportsBet2 <- tibble(
  Group = c(rep("College", 269), 
         rep("NonCollege", 214)),
  Bet = c(rep("yes", 94), 
          rep("no", 269-94),
          rep("yes", 77), 
          rep("no", 214-77))
)
glimpse(SportsBet2)

SportsBet2 %>% tabyl(Group, Bet)
```

R code for proportions test requires input as a base R `table`:
```{r}
table(SportsBet2$Group, SportsBet2$Bet)
```



## Dataset input for 2-samples proportion test (2/2)

* When using a dataset, `prop.test` requires the input `x` to be a table
* Note that we do not also specify `n` since the table already includes all needed information.

```{r}
prop.test(x = table(SportsBet2$Group, SportsBet2$Bet),
       alternative = "two.sided",
       correct = FALSE)
```


Compare output with summary stats method:

```{r}
prop.test(x = NmbrBet,                  # x is # of successes in each group 
          n = TotalNmbr,                # n is sample size in each group
          alternative = "two.sided",    # 2-sided alternative
          correct = FALSE) %>%          # no continuity correction
  tidy() %>% gt()
```


## Continuity correction: 2-prop z-test with vs. without CC

* Recall that when we approximated the 
* binomial distribution with a normal distribution to calculate a probability, 
* that we included a continuity correction (CC) 
* to account for approximating a discrete distribution with a continuous distribution.


```{r}
prop.test(x = table(SportsBet2$Group, SportsBet2$Bet),
          alternative = "two.sided", 
          correct = FALSE) %>% 
  tidy() %>% gt()
```


```{r}
prop.test(x = table(SportsBet2$Group, SportsBet2$Bet),
          alternative = "two.sided", 
          correct = TRUE) %>% 
  tidy() %>% gt()
```

Differences are small when sample sizes are large.


# Power & sample size <br> for testing proportions


## __Sample size__ calculation for testing one proportion

* Recall in our sports betting example that the null $p_0=0.36$ and the observed proportion was $\hat{p}=0.35$. 
    * The _p_-value from the hypothesis test was not significant.
    * How big would the sample size $n$ need to be in order for the _p_-value to be significant?

* [__Calculate $n$__]{style="color:green"} 
    * given $\alpha$, power ( $1-\beta$ ), "true" alternative proportion $p$, and null $p_0$:

$$n=p(1-p)\left(\frac{z_{1-\alpha/2}+z_{1-\beta}}{p-p_0}\right)^2$$

```{r}
p <- 0.35
p0 <- 0.36
alpha <- 0.05
beta <- 0.20  #power=1-beta; want >=80% power
n <- p*(1-p)*((qnorm(1-alpha/2) + qnorm(1-beta)) /
                (p-p0))^2
n
ceiling(n) 
```

_We would need a sample size of at least 17,857!_


## __Power__ calculation for testing one proportion

Conversely, we can calculate how much power we had in our example given the sample size of 269.

* [__Calculate power__]{style="color:green"}, 
    * given $\alpha$, $n$, "true" alternative proportion $p$, and null $p_0$
    
$$1-\beta=
			\Phi\left(z-z_{1-\alpha/2}\right)+\Phi\left(-z-z_{1-\alpha/2}\right)
			\quad ,\quad \text{where } z=\frac{p-p_0}{\sqrt{\frac{p(1-p)}{n}}}$$

$\Phi$ is the probability for a standard normal distribution 

```{r}
p <- 0.35; p0 <- 0.36; alpha <- 0.05; n <- 269
(z <- (p-p0)/sqrt(p*(1-p)/n))

(Power <- pnorm(z - qnorm(1-alpha/2)) +  pnorm(-z - qnorm(1-alpha/2)))
```

If the population proportion is 0.35 instead of 0.36, we only have a 6.4% chance of correctly rejecting $H_0$ when the sample size is 269.



## R package `pwr` for power analyses


* Specify all parameters _except for_ the one being solved for.

* One proportion

`pwr.p.test(h = NULL, n = NULL, sig.level = 0.05, power = NULL,`
&nbsp; &nbsp; &nbsp;    `alternative = c("two.sided","less","greater"))`

* Two proportions (same sample sizes)

`pwr.2p.test(h = NULL, n = NULL, sig.level = 0.05, power = NULL,` 
&nbsp; &nbsp; &nbsp;    `alternative = c("two.sided","less","greater"))`

* Two proportions (different sample sizes)

`pwr.2p2n.test(h = NULL, n1 = NULL, n2 = NULL, sig.level = 0.05, power = NULL,`
&nbsp; &nbsp; &nbsp;    `alternative = c("two.sided", "less","greater"))`

<hr>

$h$ is the effect size, and calculated using an arcsine transformation: 

$$h = \text{ES.h(p1, p2)} = 2\arcsin(\sqrt{p_1})-2\arcsin(\sqrt{p_2})$$


See PASS documentation for 

* [testing 1 proportion using effect size](https://www.ncss.com/wp-content/themes/ncss/pdf/Procedures/PASS/Tests_for_One_Proportion_using_Effect_Size.pdf) vs. [other ways of powering a test of 1 proportion](https://www.ncss.com/wp-content/themes/ncss/pdf/Procedures/PASS/Tests_for_One_Proportion.pdf)
* [testing 2 proportions using effect size](https://www.ncss.com/wp-content/themes/ncss/pdf/Procedures/PASS/Tests_for_Two_Proportions_using_Effect_Size.pdf) vs. [other ways of powering a test of 2 proportions](https://www.ncss.com/wp-content/themes/ncss/pdf/Procedures/PASS/Tests_for_Two_Proportions.pdf).


## `pwr`: __sample size__ for one proportion test

`pwr.p.test(h = NULL, n = NULL, sig.level = 0.05, power = NULL,`
&nbsp; &nbsp; &nbsp;    `alternative = c("two.sided","less","greater"))`

* $h$ is the effect size: `h = ES.h(p1, p2)`
    * `p1` and `p2` are the two proportions being tested
    * one of them is the null proportion $p_0$, and the other is the alternative proportion

Specify all parameters _except for_ the sample size:

```{r}
library(pwr)

p.n <- pwr.p.test(
  h = ES.h(p1 = 0.36, p2 = 0.35),
  sig.level = 0.05, 
  power = 0.80, 
  alternative = "two.sided")
p.n
```

```{r}
plot(p.n)
```


## `pwr`: __power__ for one proportion test

`pwr.p.test(h = NULL, n = NULL, sig.level = 0.05, power = NULL,`
&nbsp; &nbsp; &nbsp;    `alternative = c("two.sided","less","greater"))`


* $h$ is the effect size: `h = ES.h(p1, p2)`
    * `p1` and `p2` are the two proportions being tested
    * one of them is the null proportion $p_0$, and the other is the alternative proportion

Specify all parameters _except for_ the power:

```{r}
library(pwr)

p.power <- pwr.p.test(
  h = ES.h(p1 = 0.36, p2 = 0.35),
  sig.level = 0.05, 
  # power = 0.80, 
  n = 269,
  alternative = "two.sided")
p.power
```

```{r}
plot(p.power)
```



## `pwr`: __sample size__ for two proportions test

* Two proportions (same sample sizes)

`pwr.2p.test(h = NULL, n = NULL, sig.level = 0.05, power = NULL,` 
&nbsp; &nbsp; &nbsp;    `alternative = c("two.sided","less","greater"))`

* $h$ is the effect size: `h = ES.h(p1, p2)`;  `p1` and `p2` are the two proportions being tested

Specify all parameters _except for_ the sample size:

```{r}
p2.n <- pwr.2p.test(
  h = ES.h(p1 = 0.36, p2 = 0.35),
  sig.level = 0.05, 
  power = 0.80, 
  alternative = "two.sided")
p2.n
```

Note: $n$ in output is the __number per sample__!

```{r}
plot(p2.n)
```


## `pwr`: __power__ for two proportions test

* Two proportions (different sample sizes)

`pwr.2p2n.test(h = NULL, n1 = NULL, n2 = NULL, sig.level = 0.05, power = NULL,`
&nbsp; &nbsp; &nbsp;    `alternative = c("two.sided", "less","greater"))`


* $h$ is the effect size: `h = ES.h(p1, p2)`;  `p1` and `p2` are the two proportions being tested

Specify all parameters _except for_ the power:


```{r}
p2.n2 <- pwr.2p2n.test(
  h = ES.h(p1 = 0.36, p2 = 0.35),
  n1 = 214,
  n2 = 269,
  sig.level = 0.05, 
  # power = 0.80, 
  alternative = "two.sided")
p2.n2
```

Note: $n$ in output is the __number per sample__!

```{r}
plot(p2.n2)
```


## Where are we?

CI's and hypothesis tests for different scenarios:

$$\text{point estimate} \pm z^*(or~t^*)\cdot SE,~~\text{test stat} = \frac{\text{point estimate}-\text{null value}}{SE}$$

Day | Book | Population <br> parameter | Symbol | Point estimate | Symbol | SE
--|--|--|--|--|--|--
10 | 5.1 | Pop mean | $\mu$ | Sample mean | $\bar{x}$ | $\frac{s}{\sqrt{n}}$
10 | 5.2 | Pop mean of paired diff | $\mu_d$ or $\delta$ | Sample mean of paired diff | $\bar{x}_{d}$  | **$\frac{s_d}{\sqrt{n}}$**
11 | 5.3 | [Diff in pop <br> means]{style="color:black"} | [$\mu_1-\mu_2$]{style="color:black"} | [Diff in sample <br> means]{style="color:black"} | [$\bar{x}_1 - \bar{x}_2$]{style="color:black"}  | [**$\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$ or pooled**]{style="color:black"}
12 | 8.1 | [Pop proportion]{style="color:green"} | [$p$]{style="color:green"} | [Sample prop]{style="color:green"} | [$\widehat{p}$]{style="color:green"}  | [$\sqrt{\frac{p(1-p)}{n}}$]{style="color:red"}
12 | 8.2 | [Diff in pop <br> proportions]{style="color:green"} | [$p_1-p_2$]{style="color:green"} | [Diff in sample <br> proportions]{style="color:green"} | [$\widehat{p}_1-\widehat{p}_2$]{style="color:green"} | [$\sqrt{\frac{p_1\cdot(1-p_1)}{n_1} + \frac{p_2\cdot(1-p_2)}{n_2}}$]{style="color:red"}
