---
title: "Probability distributions in R for BSTA 511/611"
author: "Meike Niederhausen"
date: 'today'
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 3
    theme: lumen
    self-contained: true
    html-math-method: mathjax
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

* I recommend installing the `tinytex` R package to help with rendering LaTeX equations if you do not have a LaTeX program installed on your computer. 

# Binomial distribution

## Probability calculations: $P(X=x)$

### Using formula

Calculate $P(X=4)$ for a Bin(n=10, p=0.25) distribution. 

In general, for a binomial random variable, 
$$P(X=x) = \binom{n}{x}p^x q^{n-x}$$
Thus $P(X=4)$ for a Bin(n=10, p=0.25) random variable, 
$$P(X=4) = \binom{10}{4}0.25^4 0.75^{10-4} = \frac{10!}{4!(10-4)!}0.25^4 0.75^{10-4}$$
Calculate "directly" in R:
```{r}
choose(10,4) * 0.25^4 * 0.75^6

# using factorials instead of the choose function:
factorial(10)/(factorial(4)*factorial(6)) * 0.25^4 * 0.75^6
```


### Using `dbinom()`

Calculate $P(X=4)$ for a Bin(n=10, p=0.25) distribution. 

```{r}
# P(X = 4) for Bin(n=10, p=0.25) random variable
dbinom(x = 4, size = 10, prob = 0.25) 
```

Calculate $P(X=x)$ for all possible values $x$ for a Bin(n=10, p=0.25) distribution. 

```{r}
# Entire probability distribution:
# P(X = x) for all x=0,1,2,...,10 for Bin(n=10, p=0.25) random variable
dbinom(x = 0:10, size = 10, prob = 0.25) 
```

In the next section we visualize these probability distribution values.

## Histogram of probability distribution

* First, create a data frame of the distribution's possible x values and their respective probabilities:

```{r}
# Bin(n=10, p=0.25) random variable

binom_probs <- data.frame(x_values = 0:10, 
                          px = dbinom(x = 0:10, size = 10, prob = 0.25)) 
binom_probs
```

### Histogram 

The histogram below is actually created as a bar plot in ggplot:

```{r}
#| fig.height: 3
#| fig.width: 5

ggplot(binom_probs, 
       aes(x = x_values, y = px))  + 
  geom_col(fill ="cornflowerblue") +
  labs(x = "x, number of successes",
       y = "P(X=x)") +
  scale_x_continuous(breaks=0:10) +
  labs(title = "Bin(n = 10, p = 0.25) distribution")
```

## Probability calculations: $P(X \leq x)$

$$P(X\leq k) = \sum_{x=0}^{k}\binom{n}{x}p^xq^{n-x}$$

### Using formula

Calculate $P(X \leq 3)$ for a Bin(n=10, p=0.25) distribution: 

$$P(X\leq 3) = \sum_{x=0}^{3}\binom{10}{x}0.2.5^x 0.75^{10-x}$$
Calculate "directly" in R:
```{r}
# vector of x values whose probabilities need to be added
x <- 0:3  
# vector of respective binomial prob's of x values
(binom_prob_0_3 <- choose(10,x) * 0.25^x * 0.75^(10-x))
# add up the probabilities
sum(binom_prob_0_3)
```


### Using `pbinom()`

* $P(X\leq k)$ = `pbinom(q = k, size = n, prob = p, lower.tail = TRUE)`

Calculate $P(X \leq 3)$ for a Bin(n=10, p=0.25) distribution. 

```{r}
# P(X <= 3) for Bin(n=10, p=0.25) random variable
pbinom(q = 3, size = 10, prob = 0.25, lower.tail = TRUE)

# Note: setting TRUE for the lower.tail option is the default value
# This means that if we do not specify this option, 
# it will give the same result as above:
pbinom(q = 3, size = 10, prob = 0.25) 
```


## Probability calculations: $P(X \geq x)$

$$P(X\geq k) = \sum_{x=k}^{n}\binom{n}{x}p^xq^{n-x}$$

### Using formula

Calculate $P(X \geq 5)$ for a Bin(n=10, p=0.25) distribution: 

$$P(X \geq 5) = \sum_{x=5}^{10}\binom{10}{x}0.2.5^x 0.75^{10-x}$$
Calculate "directly" in R:
```{r}
# vector of x values whose probabilities need to be added
x <- 5:10  
# vector of respective binomial prob's of x values
(binom_prob_5_10 <- choose(10,x) * 0.25^x * 0.75^(10-x))
# add up the probabilities
sum(binom_prob_5_10)
```


### Using `pbinom()` with `lower.tail = TRUE`

Calculate $P(X \geq 5)$ for a Bin(n=10, p=0.25) distribution. 

$$P(X \geq 5) = 1 - P(X \leq 4) =1 - \sum_{x=0}^{4}\binom{10}{x}0.2.5^x 0.75^{10-x}$$

```{r}
# P(X >= 5) for Bin(n=10, p=0.25) random variable
# P(X >= 5) = 1 - P(X <= 4)
1 - pbinom(q = 4, size = 10, prob = 0.25, lower.tail = TRUE)
```



### Using `pbinom()` with `lower.tail = FALSE`

* $P(X \geq k) = P(X > k)$ = `pbinom(q = k, size = n, prob = p, lower.tail = FALSE)`

Calculate $P(X \geq 5)$ for a Bin(n=10, p=0.25) distribution. 

$$P(X \geq 5) = P(X > 4)$$

```{r}
# P(X >= 5) for Bin(n=10, p=0.25) random variable
# P(X >= 5) = P(X > 4)
pbinom(q = 4, size = 10, prob = 0.25, lower.tail = FALSE)
```


# Normal distribution

## Probability calculations: $P(X < x)$

$$
P(X < x) = P\Big(Z < \frac{x-\mu}{\sigma}\Big)
$$
Calculate $P(X < 10)$, for $X \sim N(\mu = 8, \sigma = 2)$.

$$
P(X < 10) = P\Big(Z < \frac{10-8}{2}\Big) = P\Big(Z < 1\Big)
$$

```{r}
# P(X < 10) without using z-scores
pnorm(q = 10, mean = 8, sd = 2, lower.tail = TRUE)

# P(X < 10) = P(Z < 1) using z-scores
pnorm(q = 1, mean = 0, sd = 1, lower.tail = TRUE)
# this works too for a standard normal distribution:
pnorm(1)
```


## Probability calculations: $P(X > x)$

$$
P(X > x) = P\Big(Z > \frac{x-\mu}{\sigma}\Big) = 1-P\Big(Z \leq  \frac{x-\mu}{\sigma}\Big)
$$
Calculate $P(X > 10)$, for $X \sim N(\mu = 8, \sigma = 2)$.

$$
P(X > 10) = P\Big(Z > \frac{10-8}{2}\Big) = 1 - P\Big(Z \leq 1\Big)
$$

```{r}
# P(X > 10) without using z-scores
1 - pnorm(q = 10, mean = 8, sd = 2, lower.tail = TRUE)
# using lower.tail = FALSE:
pnorm(q = 10, mean = 8, sd = 2, lower.tail = FALSE)

# P(X > 10) = P(Z > 1) using z-scores
1 - pnorm(q = 1, mean = 0, sd = 1, lower.tail = TRUE)
# using lower.tail = FALSE:
pnorm(q = 1, mean = 0, sd = 1, lower.tail = FALSE)
# this works too for a standard normal distribution:
pnorm(1, lower.tail = FALSE)
```


## Normal curve figure with region shaded in

* R code to create a figure of the normal distribution curve with the probability region of interest shaded in. 

Shade in the region representing $P(X>10)$, for $X \sim N(\mu = 8, \sigma = 2)$.

```{r}
#| fig.height: 3
#| fig.width: 5

# This code shades in the probability P(X > 10) 
# for X ~ N(mu = 8, sigma = 2)
# Note that I set the upper and lower bounds of the normal curve to be from mu - 4*sigma to mu + 4*sigma since these bounds include almost 100% of the area under the normal curve

mu <- 8   # specify the mean of the normal distribution
std <- 2  # specify the standard deviation of the normal distribution

# specify upper and lower bounds of shaded region below
ggplot(data.frame(x = c(mu-4*std, mu+4*std)), aes(x = x)) + 
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = std)) + 
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = std), 
          # specify the upper and lower bounds of the shaded region:
                xlim = c(10, mu+4*std),             
                geom = "area", fill = "darkblue") +
  # the breaks values below might need to be adjusted 
  # if there are too many values showing on the x-axis
  scale_x_continuous(breaks= seq(from = mu-4*std, to = mu+4*std, by= std)) +
  labs(y = "") +
  labs(title = "P(X >10) for a N(mu=8, sigma=2) distribution")
```



# Poisson distribution

## Probability calculations: $P(X=x)$

### Using formula

Calculate $P(X=3)$ for a $Pois(\lambda = 5)$ distribution. 


In general, for a Poisson random variable, 
$$P(X=x) = \frac{e^{-\lambda}\lambda^x}{x!}$$
Thus the $P(X=3)$ for a $Pois(\lambda = 5)$ random variable is 
$$P(X=3) = \frac{e^{-5}5^3}{3!}$$
Calculate "directly" in R:
```{r}
exp(-5)*(5^3)/factorial(3)
```


### Using `dpois()`

Calculate $P(X=3)$ for a $Pois(\lambda = 5)$ distribution. 

```{r}
# P(X = 3) for Pois(lambda = 5) random variable
dpois(x = 3, lambda = 5) 
```

* Calculate $P(X=x)$ for many possible values $x$ for a $Pois(\lambda = 5)$ distribution. 
* The possible values of $x$ for a Poisson distribution are $x=0, 1, 2, \ldots$, i.e., $x$ can be infinitely large.
* Below we look at probabilities just for $x=0, 1, 2, \ldots, 20$.

```{r}
# P(X = x) for x=0,1,2,...,100 for Pois(\lambda = 5) random variable
dpois(x = 0:20, lambda = 5) 

# although we didn't calculate the probabilities for all possible values of x, 
# below we see that the probabilities for the first 21 values almost add up to 1
sum(dpois(x = 0:20, lambda = 5) )
```

In the next section we visualize these probability distribution values.

## Histogram of probability distribution

* First, create a data frame of the distribution's possible x values and their respective probabilities:

```{r}
# Pois(lambda = 5) random variable

Poisson_probs <- data.frame(x_values = 0:20, 
                          px = dpois(x = 0:20, lambda = 5)) 
Poisson_probs
```

### Histogram 

The histogram below is actually created as a bar plot in ggplot:

```{r}
#| fig.height: 3
#| fig.width: 5

ggplot(Poisson_probs, 
       aes(x = x_values, y = px))  + 
  geom_col(fill ="cornflowerblue") +
  labs(x = "x, number of successes",
       y = "P(X=x)") +
  scale_x_continuous(breaks=0:20) +  # chose 20 since prob's are already tiny for this distribution
  labs(title = "Pois(lambda = 5) distribution")
```

## Probability calculations: $P(X \leq x)$

$$P(X\leq k) = \sum_{x=0}^{k}\frac{e^{-\lambda}\lambda^x}{x!}$$

### Using formula

Calculate $P(X \leq 12)$ for a $Pois(\lambda = 5)$ distribution: 

$$P(X\leq 12) = \sum_{x=0}^{12} \frac{e^{-5}5^x}{x!}$$

Calculate "directly" in R:
```{r}
# vector of x values whose probabilities need to be added
x <- 0:12  
# vector of respective Poisson prob's of x values
(Poisson_prob_0_12 <- exp(-5)*(5^x)/factorial(x))
# add up the probabilities
sum(Poisson_prob_0_12)
```


### Using `ppois()`

* $P(X\leq k)$ = `ppois(q = k, lambda = , lower.tail = TRUE)`

Calculate $P(X \leq 12)$ for a $Pois(\lambda = 5)$ distribution. 

```{r}
# P(X <= 3) for Pois(lambda = 5) random variable
ppois(q = 12, lambda = 5, lower.tail = TRUE)

# Note: setting TRUE for the lower.tail option is the default value
# This means that if we do not specify this option, 
# it will give the same result as above:
ppois(q = 12, lambda = 5) 
```


## Probability calculations: $P(X \geq x)$

$$P(X\geq k) = \sum_{x=k}^{\infty} \frac{e^{-\lambda}\lambda^x}{x!}$$

### Using formula

Calculate $P(X \geq 13)$ for a $Pois(\lambda = 5)$ distribution: 

$$P(X \geq 13) = \sum_{x=13}^{\infty} \frac{e^{-5}5^x}{x!}$$
Calculate "directly" in R:
```{r}
# vector of x values whose probabilities need to be added
x <- 13:100 # choose a big number; big enough so that the probabilities for the high x values below are tiny
# vector of respective Poisson prob's of x values
(Poisson_prob_13_100 <- exp(-5)*(5^x)/factorial(x))
# add up the probabilities
sum(Poisson_prob_13_100)
```


### Using `ppois()` with `lower.tail = TRUE`

Calculate $P(X \geq 13)$ for a $Pois(\lambda = 5)$ distribution. 

$$P(X \geq 13) = 1 - P(X \leq 12) =1 - \sum_{x=0}^{12} \frac{e^{-5}5^x}{x!}$$

```{r}
# P(X >= 13) for Pois(lambda = 5) random variable
# P(X >= 13) = 1 - P(X <= 12)
1 - ppois(q = 12, lambda = 5, lower.tail = TRUE)
```



### Using `ppois()` with `lower.tail = FALSE`

* $P(X \geq k) = P(X > k)$ = `ppois(q = k, size = n, prob = p, lower.tail = FALSE)`

Calculate $P(X \geq 13)$ for a $Pois(\lambda = 5)$ distribution. 

$$P(X \geq 13) = P(X > 12)$$

```{r}
# P(X >= 13) for Pois(lambda = 5) random variable
# P(X >= 13) = P(X > 12)
ppois(q = 12, lambda = 5, lower.tail = FALSE)
```

